---
title: "Propensity Score Matching"
subtitle: "Using GBSG2: German Breast Cancer Study Group 2 data"
author: "Tom Choi"
date: "`r Sys.Date()`"
output: 
  html_document: 
    code_folding: hide
    toc: true
    toc_depth: 5
    toc_float: 
      collapsed: true
      smooth_scroll: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(dpi = 300)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
# knitr::opts_chunk$set(fig.width=10)

```

```{r packages}
library(tidyverse)
library(flextable)
library(TH.data)
library(gtsummary)
library(MatchIt)
library(cobalt)
library(ggrepel)
library(moonBook)
library(officer)

```


## 1. Dataset overview


**Description** 

A data frame containing the observations from the GBSG2 study (n=686).

**Variables**

* `horTh`: hormonal therapy, a factor at two levels no and yes.
* `age`: age of the patients in years.
* `menostat`: menopausal status, a factor at two levels pre (premenopausal) and post (postmenopausal).
* `tsize`: tumor size (in mm).
* `tgrade`: tumor grade, a ordered factor at levels I < II < III.
* `pnodes`: number of positive nodes.
* `progrec`: progesterone receptor (in fmol).
* `estrec`: estrogen receptor (in fmol).
* `time`: recurrence free survival time (in days).
* `cens`: censoring indicator (0- censored, 1- event).

**Source**

W. Sauerbrei and P. Royston (1999). Building multivariable prognostic and diagnostic models: transformation of the predictors by using fractional polynomials. Journal of the Royal Statistics Society Series A, Volume 162(1), 71–94.


## 2. Summary table

Following is a summary table of the data set. There were significant difference in the `age` (P<0.001), `menostat` (P<0.001), and `estrec` (P=0.009) variables between hormone therapy treatments(no vs. yes).

```{r summary}

aa<-GBSG2[,1:8] %>% tbl_summary(by=horTh) %>% add_overall() %>% add_p() %>% separate_p_footnotes() %>%
    as_tibble()

dd<-GBSG2

bb <- data[,1:8] %>% tbl_summary(by=horTh) %>% add_overall() %>% add_p() %>% separate_p_footnotes() %>%
    as_tibble()
cc <- cbind(aa,bb) %>%
    flextable()
flextable(bb) %>% autofit()
# ?flextable


```


## 3. Matching using nearest neighbor 


Following is the result of 1:1 nearest neighbor matching without replacement

```{r matching}

matched <- matchit(formula = horTh ~ age + menostat + tsize + tgrade + pnodes + progrec + estrec, data = GBSG2)
matched
summary(matched)

```


## 4. Matched data (n=492)

Following is first 6 rows of the matched data.

```{r data}

data <- match.data(matched)

head(data)  %>%
    flextable() %>%
    autofit()

```

## 5. Covariates vs. Propensity Score

Following is visual inspection of whether matching is done well using a loess smoother to estimate the mean of each covariate, by treatment status, at each value of the propensity score.

```{r plot1}

plot1 <- data %>%
    mutate(across(!horTh,as.numeric)) %>%
    pivot_longer(cols = age:progrec, names_to = "variable", values_to = "response") %>%
    ggplot(aes(x = distance, y = response, color = as.factor(horTh))) +
    geom_point(alpha = 0.2, size = 1.3) +
    geom_smooth(method = "loess", se = F) +
    labs(x="Propensity score", y="", color = "horTh")+
    theme_bw() +
    facet_wrap(vars(variable), scales = "free", ncol = 2)

plot1

```


## 6. Change of Absolute Standardised Differences


```{r}
smryPSM <- function(x, show.table = TRUE, xpos = NULL, ypos = NULL) {
    res = summary(x)$sum.all[, 3]
    if (is.null(summary(x)$sum.matched)) {
        res2 = summary(x)$sum.across[, 3]
    } else {
        res2 = summary(x)$sum.matched[, 3]
    }
    res
    res2
    df1 = data.frame(value = abs(res))
    df1$name = "All Data"
    df2 = data.frame(value = abs(res2))
    df2$name = "Matched Data"
    df = rbind(df1, df2)
    rownames(df1)
    df$var = rep(rownames(df1), 2)
    df
    df3 = cbind(abs(res), abs(res2))
    colnames(df3) = c("All Data", "Matched")
    df3
    df3 = df3[order(df3[, 1], decreasing = TRUE), ]
    if (is.null(ypos)) {
        maxy = max(res)
        miny = max(res2)
        ypos = miny + (maxy - miny) * 1/2
    }
    if (is.null(xpos))
        xpos = 1.5
    table_grob = gridExtra::tableGrob(round(df3, 3), theme = gridExtra::ttheme_minimal())
    p <- ggplot(data = df, aes_string(x = "name", y = "value", color = "var", group = "var",
        label = "var")) + geom_point() + geom_line() + xlab("") + ylab("Absolute Standardized Diff in Means") +
        guides(colour = FALSE) + ggrepel::geom_text_repel(data = df[df$name == "All Data",
        ], hjust = 1.2) + ggrepel::geom_text_repel(data = df[df$name != "All Data",
        ], hjust = -0.2) + theme_bw()
    if (show.table)
        p <- p + annotation_custom(grob = table_grob, xmin = xpos, ymin = ypos)
    p
}

smryPSM(matched)
```


## 7. Distribution of Propensity Score

Following is the visualization of distribution of propensity scores (distance) of those who were matched.


```{r plot3}

plot(matched, type = "jitter", col = "blue", interactive = FALSE)

```


## 8. Histogram of Propensity Score

Following is the visualization of distribution of propensity scores (distance) before and after matching.

```{r plot4}

plot(matched, type = "hist", interactive = FALSE)

```


## 9. Balance Table

```{r table1}

bal.tab(matched, un = TRUE, stats = c("m", "v"))

```


## 10. Love plot

### 10-1. Standardized mean differences

```{r loveplot1}

love.plot(matched, stats="mean.diffs", binary="std", 
          var.order="unadjusted", grid=TRUE, thresholds=0.1,
          sample.names= c("Unmatched", "Matched"))

```

### 10-2. Variance

```{r loveplot2}

love.plot(matched, stats = "variance.ratios", 
          var.order = "unadjusted", grid=TRUE, thresholds=2, 
          sample.names= c("Unmatched", "Matched"))

```


## 11. Mirror plot


```{r mirror}

bal.plot(matched, var.name = "distance", mirror = TRUE, type = "histogram", which = "both")

```


## 12. Before and after propensity score matching


```{r}

formula_to_vars <- function(formula, allowInteraction = FALSE) {
    temp = deparse(formula)
    temp = gsub(" ", "", temp)
    temp = unlist(strsplit(temp, "~"))
    yvar = temp[1]
    yvar
    if (allowInteraction) {
        xvars = unlist(strsplit(temp[2], "+", fixed = TRUE))
    } else {
        xvars = unique(unlist(strsplit(temp[2], "[+]|[*]|:")))
    }
    xvars
    list(yvar = yvar, xvars = xvars)
}

func_pform <- function(x, digits = 3) {
    for (i in seq_along(x)) {
        if (is.na(as.numeric(x[i]))) {
            next
        } else if (as.numeric(x[i]) < 10^(-digits)) {
            x[i] = paste0("< 0.", paste0(rep("0", digits - 1), collapse = ""), "1")
        }
    }
    x
}

smry_tbl_bf_aft <- function(x, digitsstd = 3, grouplabel = NULL) {
    temp1 = formula_to_vars(x$formula)
    xvars = temp1$xvars
    yvar = temp1$yvar
    data1 = x$model$data
    if (is.null(data1)) {
        result = call2param(x$call)
        data1 = eval(parse(text = result$data))
    }
    data2 = match.data(x)[1:ncol(data1)]
    data2
    data1$matchGroup = "Before"
    data2$matchGroup = "After"
    data3 = rbind(data1, data2)
    temp = paste0("mytable(matchGroup+", yvar, "~", paste0(xvars, collapse = "+"),
        ",data=data3)")
    restable = eval(parse(text = temp))
    restable
    restable = compress(restable)
    res = list()
    form1 = paste0("%.", digitsstd, "f")
    res[[1]] = sprintf(form1, summary(x)$sum.all[, 3])
    if (is.null(summary(x)$sum.matched)) {
        res[[2]] = sprintf(form1, summary(x)$sum.across[, 3])
    } else {
        res[[2]] = sprintf(form1, summary(x)$sum.matched[, 3])
    }
    if (rownames(summary(x)$sum.all)[1] == "distance") {
        res[[1]] = res[[1]][-1]
        res[[2]] = res[[2]][-1]
    }
    start = 1
    restable[[1]]$res
    end = ncol(restable[[1]]$res) - 7
    end
    px = restable[[1]]$res$ptest
    df = list()
    for (k in 1:2) {
        df[[k]] = restable[[k]]$res[1:end]
        temp = c()
        j = 1
        for (i in 1:length(px)) {
            if (i == length(px)) {
                temp = c(temp, res[[k]][j])
                j = j + 1
            } else if ((px[i] != "") & (px[i + 1] == "")) {
                temp = c(temp, "")
            } else {
                temp = c(temp, res[[k]][j])
                j = j + 1
            }
        }
        df[[k]]$std = temp
    }
    df = cbind(df[[1]], df[[2]])
    df[[6]] <- ""
    df[[4]] <- func_pform(df[[4]])
    df[[9]] <- func_pform(df[[9]])
    df
    weighted = FALSE
    if (length(unique(match.data(x)$weights)) > 1) {
        result = getWeightedValues(x)
        df[[7]] = result$control
        df[[8]] = result$treat
        df[[9]] = result$p
        weighted = TRUE
    }
    subnames = colnames(df)
    colnames(df) = 1:10
    header1 = c("Covariates", paste0("N=", restable[[1]]$count), "", "", "", paste0("N=",
        restable[[2]]$count), "", "")
    if (is.null(grouplabel)) {
        name2 = paste0(subnames[1], "=", subnames[2:3])
    } else {
        name2 = grouplabel
    }
    header2 = c("", name2, "p", "standardized\ndifference", "", name2, "p", "standardized\ndifference")
    header3 = c("Variables", "Before Propensity Score Matching", "", "After Propensity Score Matching")
    big_border = fp_border(color = "black", width = 2)
    small_border = fp_border(color = "gray50", width = 1)
    ft <- flextable(df)
    ft
    ft <- ft %>%
        delete_part("header") %>%
        add_header_row(values = header1) %>%
        add_header_row(values = header2) %>%
        add_header_row(values = header3, colwidths = c(1, 4, 1, 4)) %>%
        hline_top(part = "header", border = big_border) %>%
        hline(i = 2, j = 2:3, border = small_border, part = "header") %>%
        hline(i = 2, j = 7:8, border = small_border, part = "header") %>%
        hline(i = 1, j = 2:5, border = small_border, part = "header") %>%
        hline(i = 1, j = 7:10, border = small_border, part = "header") %>%
        align(align = "center", part = "header") %>%
        align(align = "right", part = "body") %>%
        align(j = 1, align = "left", part = "body") %>%
        merge_at(i = 1:2, j = 1, part = "header") %>%
        merge_at(i = 2:3, j = 4, part = "header") %>%
        merge_at(i = 2:3, j = 5, part = "header") %>%
        merge_at(i = 2:3, j = 9, part = "header") %>%
        merge_at(i = 2:3, j = 10, part = "header") %>%
        hline_bottom(part = "header", border = big_border) %>%
        hline_top(part = "body", border = big_border) %>%
        hline(i = 2, j = 4:5, part = "header", border = big_border) %>%
        hline(i = 2, j = 9:10, part = "header", border = big_border) %>%
        autofit() %>%
        width(j = 6, width = 0.05) %>%
        width(j = 1, width = 0.5) %>%
        width(j = 2:5, width = 1.2) %>%
        width(j = 7:10, width = 1.2)
    if (weighted)
        ft <- footnote(ft, i = 2, j = 7:8, ref_symbols = c("a"), value = as_paragraph("Values are weighted mean $)C\016!>\017 weighted sd or weighted percentages"),
            part = "header", inline = T)
    ft
}

smry_tbl_bf_aft(matched)

```


## 13. Summary of Propensity Score Matching

We used propensity score matching to estimate the average marginal effect of the horTh on cens on those who received it accounting for confounding by the included covariates. We first attempted 1:1 nearest neighbor propensity score matching without replacement with a propensity score estimated using logistic regression of the treatment on the covariates.1:1 nearest matching on the propensity score without replacement, which yielded adequate balance, as indicated in Table and Figure. After matching, all standardized mean differences for the covariates were below 0.1 and all standardized mean differences for squares and two-way interactions between covariates were below 0.15, indicating adequate balance. The propensity score was estimated using a logistic regression of the horTh on the covariates.

